Bonding, Intimacy, and Circles of Friends: Further Computational Studies of Sulfur Dioxide Adsorption to Aqueous Surfaces

### 2011-12-05 Article.tex
New computational methods section.
Adding new sections
coordination breakdown image and more text about the coordinations.
partly finished the new coordination and temp change writeup.
Finished writing out the hand-notes for coordinations + rdfs, and the background for coordinations and cycles
Filling in the images
old?
removed line numbers
Conclusions! Finishing up the final edits before handing over to Geri.
Finishing up with edits returned from Geri. Getting ready to send out Draft 2.
Draft 3 sent to Geri, along with new abstract and title.

\documentclass{article}

\usepackage {amsmath}
\usepackage {setspace}
\usepackage [pdftex]{graphicx}
\usepackage[sort&compress]{achemso}
\usepackage{lineno}

% some formatting tags
\oddsidemargin  0.0in
\evensidemargin 0.0in
\textwidth      6.5in

\title{Temperature Effects on Surface Sulfur Dioxide}
\author{Eric S. Shamay \and Geraldine L. Richmond}

@@

%\maketitle

%\onehalfspacing
\linenumbers 
\doublespacing

%\begin{abstract}
	%...Abstract goes here...
%\end{abstract}

%\include{Introduction}
%\include{Methods}
%\include{Results}
%\include{Conclusions}

>>>>

\documentclass{article}
%\documentclass{scrartcl}

\usepackage {amsmath}
\usepackage {textcomp}
\usepackage {setspace}
\usepackage [pdftex]{graphicx}
\usepackage[sort&compress]{achemso}
%\usepackage{lineno}

% For typesetting roman numerals
\makeatletter
\newcommand{\rmnum}[1]{\romannumeral #1}
\newcommand{\Rmnum}[1]{\expandafter\@slowromancap\romannumeral #1@}
\makeatother

% some formatting tags
\usepackage[
top    = 0.5in,
bottom = 1.5in,
left   = 2.50cm,
right  = 2.50cm]{geometry}
%\oddsidemargin  0.0in
%\evensidemargin 0.0in
%\topmargin -0.5in
%\bottommargin -0.5in
%\textwidth      6.5in

\title{Bonding, Intimacy, and Circles of Friends: Further Computational Studies of Sulfur Dioxide Adsorption to Aqueous Surfaces}
\author{Eric S. Shamay \and Geraldine L. Richmond}

@@

\maketitle

%\onehalfspacing
%\linenumbers 
\doublespacing

\begin{abstract}
A water surface is a dynamic and constantly evolving terrain producing a vast array of unique molecular properties and interactions with environmental chemical species. The dynamics of water surfaces permit life on earth to continue, but also keeps us from developing a complete microscopic picture of the specific behaviors that take place within interfacial aqueous environments. This computational study examines a piece of the water puzzle by elucidating the bonding, dynamic interactions, and hydrate structures of sulfur dioxide gas adsorbing to a water surface. Results described herein address the specific ways in which sulfur dioxide gas molecules bind to a water surface, and paint a more complete picture of the adsorption pathway than was previously developed from experimental and computational studies. Ab initio molecular dynamics have been employed to study sulfur dioxide and water interactions at two environmentally relevant temperatures on a water surface. The results of this study on a common environmental and industrially important gas provide molecular insight to aid our understanding of interactions on aqueous surfaces, and gaseous adsorption processes.
\end{abstract}

\include{Introduction}
\include{Background}
\include{ComputationalMethods}
\include{Coordination-Results}
\include{Cycle-Results}
\include{Conclusions}

### 2011-12-05 Background.tex
Draft 3 sent to Geri, along with new abstract and title.

Hydrated \suldiox~clusters have been studied extensively with several recent experiments and computations forming a clearer picture of \suldiox~bulk and surface behaviors.\cite{Baer2010, Tarbuck2005, Tarbuck2006, Ota2011, Bishenden1998, Hirabayashi2006, Steudel2009, Yang2002, Hayashi1985, Moin2011, Eckl2008, Jayne1990, Jayne1990a, Donaldson1995} At a water surface, it is now known that \suldiox~forms a complex with water during adsorption, and then subsequently absorbs into the interfacial region by reaction to form ionic sulfur species.\cite{Tarbuck2005, Tarbuck2006, Ota2011} The computational study to elucidate the structure of surface hydrated \suldiox~by Baer et al. showed that the bonding coordination distributions of \suldiox~at the surface are altered relative to the bulk region, and that two coordinations dominate the distribution of bonding types: the ``SO'' and the ``SOO''. In the same work they then focused on the dominant coordination to determine the most likely cluster geometry of di- and tri-hydrate species of \suldiox. However, that study, and others probing specific hydrate structures, were performed in gas phase, under idealized conditions following geometry optimizations. None of the studies have yet focused directly on the presence of an extended hydrating structure involving \suldiox~molecules at a water surface, forming closed rings of molecular interactions. Here we apply the ab initio molecular dynamics to recreate a microscopic water surface environment. By focusing on the occurence of a specific subset of \suldiox~bonding coordination types, we probe directly a certain hydrated \suldiox~in an aqueous interfacial environment.

The cluster geometries predicted in the study by Baer et al. imply a cyclic bonding structure through the two or three hydrating waters. ``Cyclic'' here is used to denote a closed loop formed by the intermolecular hydrogen bonds, S-O interactions, and covalent bonds of the molecules involved. Figure \ref{fig:cyclic-example} depicts one such cyclic structure showing the bonds beginning on the sulfur and returning through a \suldiox-oxygen.

@@

As noted above, the optimized geometry of the \suldiox~hydrates suggests cyclic bonding structures. Geometry optimization shows the formation of these cyclic hydrate structures with two or three waters in the gas phase. A different story entirely has the potential to emerge when \suldiox~is placed in a dynamic environment such as in the course of MD simulations of an aqueous surface. Do the cyclic structures also form in the course of a dynamic bonding process on a simulated water surface, where extended hydrating structures influence \suldiox~and water behavior? To study the formation and behavior of cyclic hydrate structures we employ graph theoretical techniques on MD trajectory data. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

@@

Several arrangements of cyclic bonding structures are shown in Figure \ref{fig:cyclic-structures} for a \suldiox~molecule with three waters. Cycles with fewer or greater numbers of waters are also possible and encountered during MD. Cycle types \Rmnum{1}, \Rmnum{2}, \Rmnum{3} in Figure \ref{fig:cyclic-structures} are cyclic structures in which the \suldiox~is a member of the cycle. Types \Rmnum{4} and \Rmnum{5} do not involve the \suldiox~in the bonding cycle, but are commonly encountered as the smallest cycle types formed near the \suldiox. Type \Rmnum{3} is of particular interest because the \suldiox~in this cycle has the most frequently occuring bonding coordination (``SO'', as shown later).

>>>>

Hydrated \suldiox~clusters have been studied extensively with several recent experiments and computations forming a clearer picture of \suldiox~bulk and surface behaviors.\cite{Baer2010, Tarbuck2005, Tarbuck2006, Ota2011, Bishenden1998, Hirabayashi2006, Steudel2009, Yang2002, Hayashi1985, Moin2011, Eckl2008, Jayne1990, Jayne1990a, Donaldson1995} At a water surface, it is now known that \suldiox~forms a complex with water during adsorption, and then subsequently absorbs into the interfacial region by reaction to form ionic sulfur species.\cite{Tarbuck2005, Tarbuck2006, Ota2011} The computational study to elucidate the structure of surface hydrated \suldiox~by Baer et al. showed that the bonding coordination distributions of \suldiox~at the surface are altered relative to the bulk region, and that two coordinations dominate the distribution of bonding types: the ``SO'' and the ``SOO''. In the same work they then focused on the dominant coordination to determine the most likely cluster geometry of di- and tri-hydrate species of \suldiox. However, that study, and others probing specific hydrate structures, were performed in gas phase, under idealized conditions following geometry optimizations. None of the studies have yet focused directly on the presence of an extended hydrating structure involving \suldiox~molecules at a water surface, forming closed rings of molecular interactions. Here we apply the ab initio molecular dynamics to recreate a microscopic water surface environment. By focusing on the occurrence of a specific subset of \suldiox~bonding coordination types, we probe directly a certain hydrated \suldiox~in an aqueous interfacial environment.

The gas phase cluster geometries predicted in the study by Baer et al. imply a cyclic bonding structure through the two or three hydrating waters. ``Cyclic'' here is used to denote a closed loop formed by the intermolecular hydrogen bonds, S-O interactions, and covalent bonds of the molecules involved. Figure \ref{fig:cyclic-example} depicts one such cyclic structure showing the bonds beginning on the sulfur and returning through a \suldiox-oxygen.

@@

As noted above, the optimized geometry of the gas phase \suldiox~hydrates suggests cyclic bonding structures. Geometry optimization shows the formation of these cyclic hydrate structures with two or three waters. A different story entirely has the potential to emerge when \suldiox~is placed in a dynamic environment such as in the course of MD simulations of an aqueous surface. Do the cyclic structures also form in the course of a dynamic bonding process on a simulated water surface, where extended hydrating structures influence \suldiox~and water behavior? To study the formation and behavior of cyclic hydrate structures we employ graph theoretical techniques on MD trajectory data. Previous use of graphs in molecular computations were applied to finding stable arrangements of water clusters, ice, hydrogen bonding, extracting topological molecular properties, and cyclic structure studies.\cite{Anick2002, Huber2007, Radhakrishnan1991, Shi2005, Garcia2004, McDonald1998}

@@

Several arrangements of cyclic bonding structures are shown in Figure \ref{fig:cyclic-structures} for a \suldiox~molecule with three waters. Cycles with fewer or greater numbers of waters are also possible and encountered during MD. Cycle types \Rmnum{1}, \Rmnum{2}, \Rmnum{3} in Figure \ref{fig:cyclic-structures} are cyclic structures in which the \suldiox~is a member of the cycle. Types \Rmnum{4} and \Rmnum{5} do not involve the \suldiox~in the bonding cycle, but are commonly encountered as the smallest cycle types formed near the \suldiox. Type \Rmnum{3} is of particular interest because the \suldiox~in this cycle has the most frequently occurring bonding coordination (``SO'', as shown later).

### 2011-12-05 Coordination-Results.tex
Draft 3 sent to Geri, along with new abstract and title.

Having established the preference for an interaction through the \suldiox-sulfur atom, we look to the right side of Figure \ref{fig:bonding-coordinations} at the double-sulfur coordinations (e.g. ``SS'', ``SSO'', ``SSOO'', etc.). We see from the data that single-sulfur coordinations are overwhelmingly preferred over double-sulfur ones. Adding a third oxygen atom is also unfavorable as the ``OOO'' and ``SSOOO'' together represent less than 1\% of the trajectories, and the comparison between ``SOO'' to ``SOOO'' shows a very large decrease in occurrences.

We can now form a picture of a typical \suldiox~molecule adsorbed to a water surface across both temperatures in this study. The \suldiox~will have at least one interaction to neighboring waters through the sulfur, and will then bond asymmetrically through one of the oxygens either once or twice to water hydrogens. The \suldiox-oxygen bonds will form and break repeatedly throughout a trajectory, and overall the most dominant coordination will be the ``SO'' bonding arrangement.

@@

The binding behavior of the \suldiox~is altered by changing the temperature of the system, as evidenced in the shift in bonding coordination populations of Figure \ref{fig:bonding-coordinations} from cold to hot. In the cold temperature, the unbound, ``S'', and ``SO'' coordinations are more populated than in the hot systems. The increased temperature decreases the time spent in the unbound coordination, and causes all the coordinations to the right of ``SO'' in Figure \ref{fig:bonding-coordinations} to increase over the equivalent cold temperature populations. We see that the cold \suldiox~spends nearly four times as much time unbound as the hot \suldiox, with most of the unbound population in the hot simulations shifting to coordinations with double-oxygen and double-sulfur bonds.

@@

To verify if the ``O'' coordination forms from, and breaks quickly to the unbound coordination as is suggested by the transition frequency plots in Figure \ref{fig:coordination-transitions}, we plot the lifespans of the various coordinations in Figure \ref{fig:coordination-lifespans}. Each point in the plot represents a time during the simulation in which the \suldiox~formed the respective coordination. The vertical ``lifespan'' position is calculated directly from the amount of time spent in the given coordination before changing to another. Both cold (blue) and hot (red) data are plotted. The data of Figure \ref{fig:coordination-lifespans} show that most coordination configurations last a very brief time, with the majority forming for under 0.5 ps. The three most populous coordinations, ``S'', ``SO'', ``SOO'' (as determined from Figure \ref{fig:bonding-coordinations} by percentage) in both temperatures have coordination lifetimes of up to 1 ps, in some instances lasting up to 1.5 ps. The brevity of lifespans overall speaks to the dynamic nature of the \suldiox~surface binding. The length of time in each coordination parallels the populations of the coordinations, and suggests an ordering of steady states among various bonding coordinations. The ``unbound'' configuration stands out as an anomaly amongst the lifespans of the other bonding coordinations. The few lifespans above 1.5 ps, up to 3.25 ps long, suggest a \suldiox~that not only unbinds from the water surface, but that it also recedes far enough to avoid a quick rebinding and coordination change to ``S'' or ``O''. Those few long lived unbound data points indicate times when the \suldiox~is far from the water, residing in the gas phase until the necessary water rearrangement occurs and it is drawn back to the surface to rebind.

@@

Once the \suldiox~begins interacting with the water surface, the pathway leading back to the unbound coordination is not often traversed. As shown in Figure \ref{fig:coordination-transitions} the dominant coordination transitions occur between the ``S-SO-SOO'' group of coordinations. This suggests that the \suldiox-sulfur interaction to a water oxygen has a much longer lifespan than the \suldiox-oxygen bonding to water hydrogens. The difference between the ``S-SO-SOO'' coordinations is an addition or removal of oxygen bonds, and the frequent transitions between them show that the bonds to \suldiox-oxygens are quickly forming and breaking. For the \suldiox-sulfur interaction to break, the \suldiox~must enter a non-sulfur coordination (e.g. ``O'', ``OO'', unbound, etc.) or a coordination with more than a single sulfur interaction (e.g. ``SS'', ``SSO'', etc.). The transitions to coordinations that allow for breaking of the \suldiox-sulfur interactions, or switching the interaction to another water, are infrequent compared to those leading to an oxygen bond transition. Thus, the \suldiox~spends most of its time while bound to the surface waters breaking and forming hydrogen-bonds through its oxygens, and interacting with neighboring waters through a more persistent interaction via the \suldiox-sulfur atom.

>>>>

Having established the preference for an interaction through the \suldiox-sulfur atom, we look to the right side of Figure \ref{fig:bonding-coordinations} at the double-sulfur coordinations (e.g. ``SS'', ``SSO'', ``SSOO'', etc.). From the data it is clear that single-sulfur coordinations are overwhelmingly preferred over double-sulfur ones. Adding a third oxygen atom is also unfavorable as the ``OOO'' and ``SSOOO'' together represent less than 1\% of the trajectories, and the comparison between ``SOO'' to ``SOOO'' shows a very large decrease in occurrences.

A picture can now be formed of a typical \suldiox~molecule adsorbed to a water surface across both temperatures in this study. The \suldiox~will have at least one interaction to neighboring waters through the sulfur, and will then bond asymmetrically through one of the oxygens either once or twice to water hydrogens. The \suldiox-oxygen bonds will form and break repeatedly throughout a trajectory, and overall the most dominant coordination will be the ``SO'' bonding arrangement.

@@

The binding behavior of the \suldiox~is altered by changing the temperature of the system, as evidenced in the shift in bonding coordination populations of Figure \ref{fig:bonding-coordinations} from cold to hot. In the cold temperature, the unbound, ``S'', and ``SO'' coordinations are more populated than in the hot systems. The increased temperature decreases the time spent in the unbound coordination, and causes all the coordinations to the right of ``SO'' in Figure \ref{fig:bonding-coordinations} to increase over the equivalent cold temperature populations. As shown, the cold \suldiox~spends nearly four times as much time unbound as the hot \suldiox, with most of the unbound population in the hot simulations shifting to coordinations with double-oxygen and double-sulfur bonds.

@@

To verify if the ``O'' coordination forms from, and breaks quickly to the unbound coordination as is suggested by the transition frequency plots in Figure \ref{fig:coordination-transitions}, the lifespans of the various coordinations are plotted in Figure \ref{fig:coordination-lifespans}. Each point in the plot represents a time during the simulation in which the \suldiox~formed the respective coordination. The vertical ``lifespan'' position is calculated directly from the amount of time spent in the given coordination before changing to another. Both cold (blue) and hot (red) data are plotted. The data of Figure \ref{fig:coordination-lifespans} show that most coordination configurations last a very brief time, with the majority forming for under 0.5 ps. The three most populous coordinations, ``S'', ``SO'', ``SOO'' (as determined from Figure \ref{fig:bonding-coordinations} by percentage) in both temperatures have coordination lifetimes of up to 1 ps, in some instances lasting up to 1.5 ps. The brevity of lifespans overall speaks to the dynamic nature of the \suldiox~surface binding. The length of time in each coordination parallels the populations of the coordinations, and suggests an ordering of steady states among various bonding coordinations. The ``unbound'' configuration stands out as an anomaly amongst the lifespans of the other bonding coordinations. The few lifespans above 1.5 ps, up to 3.25 ps long, suggest a \suldiox~that not only unbinds from the water surface, but that it also recedes far enough to avoid a quick rebinding and coordination change to ``S'' or ``O''. Those data points of a few long lived unbound species indicate times when the \suldiox~is far from the water, residing in the gas phase until the necessary water rearrangement occurs and it is drawn back to the surface to rebind.

@@

Once the \suldiox~begins interacting with the water surface, the pathway leading back to the unbound coordination is not often traversed. As shown in Figure \ref{fig:coordination-transitions} the dominant coordination transitions occur between the ``S-SO-SOO'' group of coordinations. This suggests that the \suldiox-sulfur interaction to a water oxygen has a much longer lifespan than the \suldiox-oxygen bonding to water hydrogens. The difference between the ``S-SO-SOO'' coordinations is an addition or removal of oxygen bonds. The frequent transitions between them show that the bonds to \suldiox-oxygens are quickly forming and breaking. For the \suldiox-sulfur interaction to break, the \suldiox~must enter a non-sulfur coordination (e.g. ``O'', ``OO'', unbound, etc.) or a coordination with more than a single sulfur interaction (e.g. ``SS'', ``SSO'', etc.). The transitions to coordinations that allow for breaking of the \suldiox-sulfur interactions, or switching the interaction to another water, are infrequent compared to those leading to an oxygen bond transition. Thus, the \suldiox~spends most of its time while bound to the surface waters breaking and forming hydrogen-bonds through its oxygens, and interacting with neighboring waters through a more persistent interaction via the \suldiox-sulfur atom.

### 2011-12-05 Introduction.tex
Draft 3 sent to Geri, along with new abstract and title.

The molecular nature of adsorption of gas molecules onto a water surface is one of the remaining largely uncharted territories of surface chemistry. Although gas uptake into aqueous systems occurs often environmentally and industrially, we still know very little about the process and the details of the adsorption reactions, and certainly less than what we know about gaseous adsorption on a solid surface. How does gas initially bind to a water surface, and what steps are involved in the subsequent adsorption? How does an unbound gas molecule near a water surface affect the water to which it will bind? What is the structure of hydrating waters in the surface region, and how does a hydrated solute molecule behave differently than as a gas? Experiments to address these questions provide valuable information, but have to date never fully characterize microscopic events and behaviors. However, these systems can be fully characterized computationally, and when coupled to the previous experimental work can provide a much more complete picture of gaseous adsorption to aqueous surfaces.

>>>>

The molecular nature of the adsorption of gas molecules onto a water surface is one of the remaining largely uncharted territories of surface chemistry. Although gas uptake into aqueous systems occurs often environmentally and industrially, we still know very little about the process and the details of the adsorption reactions, and certainly less than what we know about gaseous adsorption on a solid surface. How does a gas initially bind to a water surface, and what steps are involved in the subsequent adsorption? How does an unbound gas molecule near a water surface affect the water to which it will bind? What is the structure of hydrating waters in the surface region, and how does a hydrated solute molecule behave differently than as a gas? Experiments to address these questions provide valuable information, but have to date never fully characterize microscopic events and behaviors. However, these systems can be fully characterized computationally, and when coupled to the previous experimental work can provide a much more complete picture of gaseous adsorption to aqueous surfaces.
